Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-12-28 manuscript/manuscript.tex
Minor fixes to the adaptive discretization text
Mostly moved the definition of the tesseroid size to Step 1 so that it
follows the definition of the inequality. It was breaking the flow of
the text before that.

\citet{Ku1977} noticed that the GLQ integration
becomes less accurate when the computation point is closer to the
mass element.
One way to prevent this from happening would be to increase the GLQ order.
Doing so would uniformly increase the number of point masses inside the
tesseroid volume.
However, an increase in the point mass concentration is only required close to the
computation point \citep{Uieda2016}.
Alternatively, \citet{Li2011} proposed an adaptive
discretization algorithm which keeps the GLQ order fixed and divides the
tesseroid based on a ratio between the distance to the computation
point and its dimensions.
This algorithm produces a more efficient computation because an increased concentration
of point masses is produced where it is needed more.
\citet{Uieda2016} developed a modified version of this algorithm that speeds the
computation of the distances.
Both algorithms (the original one from \citet{Li2011} and the modified
version of \citet{Uieda2016}) perform tesseroid subdivisions in the latitudinal,
longitudinal and radial directions, thus we can define them as three dimensional
adaptive discretization algorithms.
On the other hand, \citet{Lin2018} proposed a two dimensional discretization algorithm
that subdivides the tesseroid only on the latitudinal and longitudinal directions.
Removing a dimension from the splitting makes the computation more efficient, while
keeping an acceptable accuracy on the gravitational fields \citep{Lin2018}.

Here we will use a two dimensional version of the adaptive discretization algorithm
developed by \citet{Uieda2016}: the distances from the tesseroid and the computation
point are computed as \citet{Uieda2016} do, but the tesseroid will only be split on the
longitudinal and latitudinal directions.
What follows is a summary of the algorithm and the reader is referred to
\citet{Uieda2016} for a detailed description.

Lets consider a computation point located at $(r, \phi, \lambda)$ and an arbitrary
tesseroid defined by the boundaries $r_1$, $r_2$, $\phi_1$, $\phi_2$, $\lambda_1$ and
$\lambda_2$, which geometric centre is located at $(r_t, \phi_t, \lambda_t)$, where

\begin{equation}
    r_t = \frac{r_2 + r_1}{2}, \quad
    \phi_t = \frac{\phi_2 + \phi_1}{2}, \quad
    \lambda_t = \frac{\lambda_2 + \lambda_1}{2}.
\end{equation}

We can define the longitudinal ($L_\lambda$) and latitudinal ($L_\phi$) dimensions of
the tesseroid as follows:

\begin{equation}
    L_\lambda = r_2 \arccos(\sin^2\phi_t +
        \cos^2\phi_t\cos(\lambda_2 - \lambda_1)),
    \label{eq:sizelon}
\end{equation}

\textit{Step 1}: Check that the tesseroid satisfies the following inequality for each
dimension $L_i (i \in \{\lambda, \phi\})$ of the tesseroid:

@@

\noindent
in which $D$ is a positive scalar called the distance-size ratio and $d$ is the distance
between the computation point and the geometric centre of the tesseroid

@@

@@

\textit{Step 3}:
If any dimension fails inequality~\ref{eq:condition}, split the tesseroid in half along
the offending dimensions.
Repeat steps 1-3 for all smaller tesseroids until none are left.

\textit{Final step}:
By the end of the algorithm, a second-order GLQ will be applied on each small tesseroid
to compute the gravitational field they generate, and the sum of these results will be
the gravitational effect of the original tesseroid.

Notice that the distance-size ratio $D$ determines how many
times the tesseroids will be divided.
Therefore, it effectively regulates both the accuracy of the algorithm and its
computation time.
An optimal value for $D$ cannot be directly calculated from the desired accuracy level.
Instead, it is empirically determined by comparing the numerical results with the
analytical solution for a spherical shell.
\citet{Uieda2016} used a shell with homogeneous density to determine optimal values of
$D$.
Here, we will repeat the numerical experiment using analytical expressions for shells
with density varying according to certain functions of $r$.
This numerical experiment will also show that the two dimensional adaptive
discretization algorithm can achieve the same accuracy as the original three dimensional
version, though it reduces the number of tesseroid subdivisions and thus the computation
time.

>>>>

\citet{Ku1977} noticed that the GLQ integration
becomes less accurate when the computation point is closer to the
mass element.
One way to prevent this from happening would be to increase the GLQ order.
Doing so would uniformly increase the number of point masses inside the
tesseroid volume.
However, an increase in the point mass concentration is only required close to the
computation point \citep{Uieda2016}.
Alternatively, \citet{Li2011} proposed an adaptive
discretization algorithm which keeps the GLQ order fixed and divides the
tesseroid based on a ratio between the distance to the computation
point and its dimensions.
This algorithm produces a more efficient computation because an increased concentration
of point masses is produced where it is needed more.
\citet{Uieda2016} developed a modified version of this algorithm along with an efficient
computational implementation.
Both algorithms perform tesseroid subdivisions in the latitudinal,
longitudinal and radial directions, thus we can define them as three-dimensional
adaptive discretization algorithms.
On the other hand, \citet{Lin2018} proposed a two dimensional discretization algorithm
that subdivides the tesseroid only on the latitudinal and longitudinal directions.
Removing a dimension from the discretization makes the computation more efficient by
reducing the number of tesseroids in the model, while
retaining an acceptable accuracy \citep{Lin2018}.

Here we will follow \citet{Lin2018} and use a two dimensional version of the adaptive
discretization of \citet{Uieda2016}.
What follows is a summary of the algorithm and the reader is referred to
\citet{Uieda2016} for a detailed description.

\textit{Step 1}: Check that the tesseroid satisfies the following inequality for each
dimension $L_i\ (i \in \{\lambda, \phi\})$ of the tesseroid:

@@

\noindent
in which $D$ is a positive scalar called the distance-size ratio, $d$ is the distance
between the computation point and the geometric centre of the tesseroid

@@

\begin{equation}
    r_t = \frac{r_2 + r_1}{2}, \quad
    \phi_t = \frac{\phi_2 + \phi_1}{2}, \quad
    \lambda_t = \frac{\lambda_2 + \lambda_1}{2}.
\end{equation}
\\
\noindent
and the dimensions of the tesseroid are defined as

\begin{equation}
    L_\lambda = r_2 \arccos(\sin^2\phi_t +
        \cos^2\phi_t\cos(\lambda_2 - \lambda_1)),
    \label{eq:sizelon}
\end{equation}
\\
\noindent and

@@

\textit{Step 3}:
If inequality~\ref{eq:condition} does not hold for one of the dimensions
(longitudinal or latitudinal), split the tesseroid in half along that dimension.
Repeat steps 1-3 for all smaller tesseroids until none are left that violate
inequality~\ref{eq:condition}.

\textit{Final step}:
By the end of the algorithm, a second-order GLQ will have been applied to each smaller
tesseroid and the running total gathered in step 2 will be the gravitational effect of
the original tesseroid.

The distance-size ratio $D$ determines how many times the tesseroids will be divided.
Therefore, it effectively regulates both the accuracy of the algorithm and its
computation time.
An optimal value for $D$ cannot be directly calculated from the desired accuracy level.
Instead, it is empirically determined by comparing the numerical results with the
analytical solution for a spherical shell.
\citet{Uieda2016} used a shell with homogeneous density to determine optimal values of
$D$.
Here, we will repeat the numerical experiment using analytical expressions for shells
with density varying according to functions of $r$.
This experiment will also test whether the same values of $D$ determined by
\citet{Uieda2016} can be used for the two dimensional adaptive discretization.
